How predictable is extinction?

### 2018-12-07 doc/intro.tex
growing introduction

How predictable is extinction? Forecasting species survival at million-year timescales.

One of the great promises of paleobiology is that by studying the past we can better predict the future.  This promise is particularly pertinent given as risk assessments for some modern species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.  Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)? (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?

To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). Using a model of species survival, we analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature. Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future.

The effects of the historical covariates are extremely small and vary considerably over time. For example, the effect of change in geographic range size can either increase or decrease probability of extinction depending on when in the Cenozoic the observation takes place. The improvement in predictive power by including these historical covariates is modest at best and reflects the rarity of extinction events (i.e. class-imbalance) and the extremely stochastic nature of species survival. Correcting for class imbalance, we find that in-sample model performance measures are approximately equal to out-of-sample performance as estimated from cross-validation. These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small.  This result implies that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.

Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. The greatest source of variance in survival probability is the timing of observation. Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age.

Ultimately, we find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. However, the effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.

Instead of modeling geographic range through time, we instead model extinction probability as a function of change in geographic range. This formulation means that, if there if geographic range trajectory is deterministic and not entirely Markovian, that past geographic range, specifically the change in geographic range, and not just current geographic range predicts the probability of extinction. For example, is extinction probability affected by a species has been decreasing in geographic range over time or does it just matter if that species has below average geographic range regardless of how it got there?

Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My scale. Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales. Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record. By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.

>>>>

Paleobiology is concerned with understanding why certain species go extinct while others do not CITATIOn. Being able to predict which species are more likely to go extinct than others is critical for making good conservation decisions to limit the impact of the current biodiversity crisis. We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable. One way of approaching this problem is the analyze the past and use that to predict the future. The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant. By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions.

By studying how species vary in their extinction risk over time and we can assess which species are at greater risk under unobserved conditions. We that that extinction risk varies over time in both intensity (average rate) and selectivity (difference in risk between taxa).

By including and modeling this natural variation in extinction risk, we are able to improve our overall predictions because we incorporate and explictly model differences between observations from across a range of possible intensities and selectivies

or the intrisic risk of extinction -- all species can go extinct at any ``moment,'' what are the relative risks between groups and over time
if we gather the kinds of predictors that apparently matter, and use them to make specific predictions of extinction 
If we gather the kinds of predictors that have been found to correlate strongly with extinction risk, how well can we predict species extinction?

Past studies have focused on identifying and measuring the effect of various predictors on extinction risk e.g. does difference in diet or geographic range affect differences in extinction risk?
Here we focus on the quality of our predictions e.g. how accurately can we predict if one species is more likely to go extinct than another?
increasing focus on prediction -- how accurate are we?
understanding versus prediction
long history focused on what determinates are important, but not on how accurate our predictions are given these factors.
the evaluate how well we make predictions in the present
trajectories studied before as phenonmenon, but not as tool -- we are using this information as tool!

has been suggested CITE seth wolfgang michael
incorporating extinction survival data from the fossil record can aide in predicting extinction risk of extant species -- the present must at some level be a function of the past
what has not been evaluated is given selectivity changes and intensity changes, how accurate are assessements based on the past likely to be in the future.

for this kind of excercise the best resolved fossil record of skeletonihze marine plantonic microorganisms (names). high temporal resolution, lots of samples, internally consistent taxonomy. 
Rich fossil record -- we might be able to better understand predicting extinction 
trajectories studied before as phenonmenon, but not as tool -- we are using this information as tool!
To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). Using a model of species survival, we analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature. 
cite the papers that evalutae the importance of the predcitors.
60 million year record lets us ask the 3 main things -- accurate, how much variation in accuracy,

there have been major climate and oceanographic events that have prfoundly influenced the macroevolution of these groups CITATIONS.
intervals of time that are particularly 
things have changed over time -- these moments of big change are ``different'' so predicting them well is porentially important

the amoung of predictive power that the past has at any given 
just the last million years may matter -- so change to range matters
predictions to be better when similar stress over long period of time -- better on the average
a more protracted event/transition (Eo-Mi) -- might expect there to be more predictive power then because past might matter more (key is protraction)

\end{document}
